Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-04 paper/paper.tex
notes removal
fix apw name and affil
serious abstract changes

\author[0000-0003-0872-7098]{Adrian Price-Whelan}
\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

\begin{abstract}\noindent % trust me
Conversion of a noisy parallax measurement into a posterior
belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the
\gaia\ \tgas\ Catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information for $1.2$ million \tgas\ stars
from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a
transformed space where the measurement uncertainty is close to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the
\tgas\ catalog entries by a median factor of 1.2, and 14 percent
are more precise by a factor $>2$. They are also more precise than
previous examples of Bayesian distance estimates using a spatial prior.
We independently validate our parallax inferences distances by looking
at members of Milky Way star cluster M67, which is not visible as a
cluster in the \tgas\ parallax estimates, but appears clearly as a
cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for 1.4 million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

\keywords{
  catalogs
  ---
  Hertzsprung--Russell and C--M diagrams
  ---
  methods: statistical
  ---
  parallaxes
  ---
  stars: distances
  ---
  stars: statistics
}

@@

%How this compares to other Bayesian attempts at inferring distances,
%like, eg, the Tri--Coryn papers. Favorably! And with \emph{fewer}
%assumptions, in fact! What form should this comparison be in?
There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth.
It is a fairly weak prior.
It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax.
A comparison of the posteriors using the two different priors, our \cmd\ prior and the EDSD, is shown in Figure \ref{fig:comparePrior}.
Here the parallax posterior pdf expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.
%LMA: chat about bias?

>>>>

\author[0000-0003-0872-7098]{Adrian M. Price-Whelan}
\affil{Department of Astrophysical Sciences, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.2$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This data-driven model contains no knowledge of stellar astrophysics or the
Milky Way, but rather derives its precision from its generative model of noisy
parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The \xd-optimized prior is used to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a posterior parallax and distance sampling for 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

\keywords{
  catalogs
  ---
  Hertzsprung--Russell and C--M diagrams
  ---
  methods: statistical
  ---
  parallaxes
  ---
  stars: distances
}

@@

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth.
It is a fairly weak prior.
It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax.
A comparison of the posteriors using the two different priors, our \cmd\ prior and the EDSD, is shown in Figure \ref{fig:comparePrior}.
Here the parallax posterior pdf expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.
